---
title: "Reading Log"
subtitle: "A list of what I am currently reading, have read, and plan to read."
date-modified: 10/4/2025
date-format: long
echo: false
---

I have been focusing on reading important philosophical and psychological texts as well as textbooks pertaining to R, machine learning, and data science.

I also listen to podcasts and lectures to center my mind on the important things in life.
Centering around being better disciplined and having mental fortitude alongside physical strength and intelligence.

During the academic year my reading frequency will probably slow to make room for assigned readings.

# Currently Reading

```{r}
divs <- c(
  bible = 279 / 365, # days
  wwwwg = 260 / 505, # pages
  econ  = 345 / 634, # pages
  add   =  57 / 243
) |> 
  scales::label_percent(0.01)() |> 
  purrr::map(
    ~ glue::glue(
      '<div class="progress"><div class="progress-bar" role="progressbar" style="width: {.x};" aria-valuenow="50" aria-valuemin="0" aria-valuemax="100"></div></div>'
    )
  )
```

*The Bible* (ESV, 2016) - Read chronologically.

```{r}
#| results: asis
divs$bible
```

---

[*We Who Wrestle With God: Perceptions of the Divine*](https://www.penguinrandomhouse.com/books/714076/we-who-wrestle-with-god-by-jordan-b-peterson/) (2024) by Jordan Peterson.

```{r}
#| results: asis
divs$wwwwg
```

---

[*Basic Economics*](https://en.wikipedia.org/wiki/Basic_Economics) (2000, 5th edition) by [Thomas Sowell](https://en.wikipedia.org/wiki/Thomas_Sowell).

```{r}
#| results: asis
divs$econ
```

---

[*Angels, Demons, and the Devil*](https://www.amazon.com/Angels-Demons-and-the-Devil/dp/0966006062) (2013) by F. LaGard Smith

```{r}
#| results: asis
divs$add
```

---

# Future Reading

These are books which are next in line to read.

1. [*The Return of the King*](https://en.wikipedia.org/wiki/The_Return_of_the_King) (1994 edition) by J. R. R. Tolkien.

2. [*The American Story: Building the Republic*](https://www.goodreads.com/book/show/208562415-the-american-story) (2024) by David and Tim Barton.

# Podcasts/Lectures

- [*The Jordan B. Peterson Podcast*](https://open.spotify.com/show/1Zw2DKjelPnuEYpydFlhgN?si=fd1d1b72d8464e06) by [Jordan Peterson](https://www.jordanbpeterson.com/).

- [*The Impossible Life*](https://open.spotify.com/show/3McmSdhO7OR6GB97Nqq69I?si=344f35bad6284a28) by [Garrett Unclebach & Nick Surface](https://www.theimpossible.life/).

- [*Huberman Lab*](https://open.spotify.com/show/79CkJF3UJTHFV8Dse3Oy0P?si=ba510b7eac074442) by Andrew Huberman.

# History

## 2025

My goal for this year is 12 books.

```{r}
#| results: asis
goal <- scales::label_percent(0.01)(9 / 12)

glue::glue(
    '<div class="progress"><div class="progress-bar" role="progressbar" style="width: {goal}; background-color:#d9534f;" aria-valuenow="50" aria-valuemin="0" aria-valuemax="100"></div></div>'
)
```
---

- [*Philosophy and Human Geography*](https://archive.org/details/philosophyhumang00rjjo) (1986, 2nd edition) by [Ron Johnston](https://en.wikipedia.org/wiki/Ron_Johnston_(geographer)).

- [*The Screwtape Letters*](https://en.wikipedia.org/wiki/The_Screwtape_Letters) (1942) by C. S. Lewis.

- [*Advanced R*](https://adv-r.hadley.nz/index.html) (2019, 2nd edition) by Hadley Wickham.

- [*The Two Towers*](https://en.wikipedia.org/wiki/The_Two_Towers) (1994 edition) by J. R. R. Tolkien.

- [*Start, Stop, or Grow?*](https://robertgrayatkins.com/book/) (2022) by [Robert Gray Atkins](https://robertgrayatkins.com/).

- [*Tidy Modeling with R*](https://www.tmwr.org/) (2023) by Max Kuhn and Julia Silge.

- [*The Fellowship of the Ring*](https://en.wikipedia.org/wiki/The_Fellowship_of_the_Ring) (1994 edition) by [J. R. R. Tolkien](https://en.wikipedia.org/wiki/J._R._R._Tolkien).

- [*Hands-On Machine Learning with R*](https://bradleyboehmke.github.io/HOML/) (2020) by Bradley Boehmke and Brandon Greenwell.

- [*Mere Christianity*](https://en.wikipedia.org/wiki/Mere_Christianity) (1952) by [C. S. Lewis](https://en.wikipedia.org/wiki/C._S._Lewis).

## 2024

I started reading more deliberately in the fall.

- *The Bible* - Most of the Old Testament I read in the NIV, but the rest I read in the ESV.

- [*Awesome Encounters with God: The Miracles of Jesus*](https://www.thriftbooks.com/w/awesome-encounters-with-god---the-miracles-of-jesus_dr-russell-bone/52477651/#edition=71087318) (2003) by Dr. Russel Bone.

- [*ggplot2: Elegant Graphics for Data Analysis*](https://ggplot2-book.org/) (3rd edition) by Wickham, Navarro and Pedersen.

- [*Hillbilly Elegy*](https://en.wikipedia.org/wiki/Hillbilly_Elegy) (2016) by Vice President [J. D. Vance](https://en.wikipedia.org/wiki/JD_Vance).

- [*R for Data Science*](https://r4ds.hadley.nz/) (2017, 2nd edition) by [Hadley Wickham](https://github.com/hadley).

- [*The American Story: The Beginnings*](https://www.google.com/books/edition/The_American_Story/r3GVzQEACAAJ?hl=en) (2020) by David and Tim Barton.

- [*The Hobbit*](https://en.wikipedia.org/wiki/The_Hobbit) (1995 edition) by J. R. R. Tolkien.
