---
title: "Reading Log"
subtitle: "A list of what I am currently reading, have read, and plan to read."
date-modified: 4/5/2025
date-format: long
echo: false
---

I have been focusing on reading important philosophical and psychological texts as well as textbooks pertaining to R, machine learning and data science.

I also listen to podcasts and lectures to center my mind on the important things in life.
Centering around being better disciplined and having mental fortitude alongside physical strength and intelligence.

# Currently Reading

```{r}
divs <- c(
  bible = 95  / 365, # days
  advr  =  6  / 25,  # chapters
  lotr  = (452 - 403) / (725 - 403), # pages
  wwwwg = 74  / 505, # pages
  ssog  = 62  / 159  # pages
) |> 
  scales::label_percent(0.01)() |> 
  purrr::map(
    ~ glue::glue(
      '<div class="progress"><div class="progress-bar" role="progressbar" style="width: {.x};" aria-valuenow="50" aria-valuemin="0" aria-valuemax="100"></div></div>'
    )
  )
```

*The Bible* - I am tackling the Bible in one year, read chronologically. I hope to continue in future years with the same or similar reading plan. I am also taking notes whenever my schedule permits.

```{r}
#| results: asis
divs$bible
```

---

[*Advanced R*](https://adv-r.hadley.nz/index.html) (2nd edition, 2019) by Hadley Wickham.

```{r}
#| results: asis
divs$advr
```

---

[*The Two Towers*](https://en.wikipedia.org/wiki/The_Two_Towers) (1994 edition) by J. R. R. Tolkien.

```{r}
#| results: asis
divs$lotr
```

---

[*We Who Wrestle With God: Perceptions of the Divine*](https://www.penguinrandomhouse.com/books/714076/we-who-wrestle-with-god-by-jordan-b-peterson/) (2024) by Jordan Peterson.

```{r}
#| results: asis
divs$wwwwg
```

---

[*Start, Stop, or Grow?*](https://robertgrayatkins.com/book/) (2022) by [Robert Gray Atkins](https://robertgrayatkins.com/).

```{r}
#| results: asis
divs$ssog
```

---

# Future Reading

These are books that are next in line to read.

1. [*The Screwtape Letters*](https://en.wikipedia.org/wiki/The_Screwtape_Letters) (1942) by C. S. Lewis.

2. [*The Return of the King*](https://en.wikipedia.org/wiki/The_Return_of_the_King) (1994 edition) by J. R. R. Tolkien.

Plus three more books to reach my goal for the year.

# Podcasts/Lectures

- [*The Jordan B. Peterson Podcast*](https://open.spotify.com/show/1Zw2DKjelPnuEYpydFlhgN?si=fd1d1b72d8464e06) by [Jordan Peterson](https://www.jordanbpeterson.com/).

- [*The Impossible Life*](https://open.spotify.com/show/3McmSdhO7OR6GB97Nqq69I?si=344f35bad6284a28) by [Garrett Unclebach & Nick Surface](https://www.theimpossible.life/).

- [*James B. Crichton Lectures*](https://open.spotify.com/show/4mgsxnDBgwKIrgn7Z4Wkya?si=5b4e31ed13b4403a) by James Crichton.

# History

## 2025

My goal for this year is 12 books.

<div class="progress"><div class="progress-bar" role="progressbar" style="width: 33.33%; background-color:#d9534f;" aria-valuenow="50" aria-valuemin="0" aria-valuemax="100"></div></div>

---

- [*Tidy Modeling with R*](https://www.tmwr.org/) (2023) by Max Kuhn and Julia Silge.

- [*The Fellowship of the Ring*](https://en.wikipedia.org/wiki/The_Fellowship_of_the_Ring) (1994 edition) by [J. R. R. Tolkien](https://en.wikipedia.org/wiki/J._R._R._Tolkien).

- [*Hands-On Machine Learning with R*](https://bradleyboehmke.github.io/HOML/) (2020) by Bradley Boehmke and Brandon Greenwell.

- [*Mere Christianity*](https://en.wikipedia.org/wiki/Mere_Christianity) (1952) by [C. S. Lewis](https://en.wikipedia.org/wiki/C._S._Lewis).

## 2024

I started reading more deliberately in the fall.

- *The Bible* - Most of the Old Testament I read in the NIV, but the rest I read in the ESV.

- [*Awesome Encounters with God: The Miracles of Jesus*](https://www.thriftbooks.com/w/awesome-encounters-with-god---the-miracles-of-jesus_dr-russell-bone/52477651/#edition=71087318) (2003) by Dr. Russel Bone.

- [*ggplot2: Elegant Graphics for Data Analysis*](https://ggplot2-book.org/) (3rd edition) by Wickham, Navarro and Pedersen.

- [*Hillbilly Elegy*](https://en.wikipedia.org/wiki/Hillbilly_Elegy) (2016) by Vice President [J. D. Vance](https://en.wikipedia.org/wiki/JD_Vance).

- [*R for Data Science*](https://r4ds.hadley.nz/) (2nd edition, 2017) by [Hadley Wickham](https://github.com/hadley).

- [*The American Story: The Beginnings*](https://www.google.com/books/edition/The_American_Story/r3GVzQEACAAJ?hl=en) (2020) by David and Tim Barton.

- [*The Hobbit*](https://en.wikipedia.org/wiki/The_Hobbit) (1995 edition) by J. R. R. Tolkien.
